A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-06-10 jacobian_calc_CMAME.tex
Fix editorial comments and grammatical corrections

%Package for adding notes/comments/and tracking changes
\usepackage[color=red!40, textsize=scriptsize]{todonotes}

@@

    In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an underexplored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the  method with other techniques for numerical derivative calculation: automatic differentiation, forward finite difference, and central finite difference. The complex-step method was implemented in a massively parallel computational peridynamics code for the purpose of this comparison.  The methods were compared through profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code.  Additionally, motivated by the reproducible research movement, all of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Newton's method \sep Newton-Raphson \sep numerical differentiation \sep complex-step \sep finite difference \sep automatic differentiation \sep tangent-stiffness \sep tangent-stiffness
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%\MSC [2010] 65D25 
\end{keyword}

@@

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent\todo[]{Re: II.1}\footnote{In the context of plasticity thoery, the algorithmic consistency of the tangent-stiffness matrix refers to consistency with the integration of a plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.} tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

@@

The scope of the application component of this study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the  \todo[]{Re: I.1} complex-step method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of forward difference, central difference, and automatic-differentiation methods needed for comparison to the complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as those required by the study.

The aim of this paper is not only to \todo[]{Re: I.1} study the complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: In Section~\ref{subsec:TS} a description of tangent-stiffness matrices, detailed directions for producing tangent-stiffness matrices with each of the methods identified above, and a description and justification of the  \todo[]{Re: I.1} complex-step method for calculating tangent-stiffness matrices, in Section~\ref{subsubsec:Impl} a description of implementing the complex-step method in the \emph{Peridigm} software, in Section~\ref{JGAM} a description of the quantities-of-interest used to rank the methods, in Section~\ref{sec:Res} a presentation and analysis of the results of the comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.  Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.  The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion.  \todo[]{Re: II.5}This open source resource is available at the corresponding author's website or at \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

This section contains background information on the differentiation techniques underlying the tangent-stiffness matrix calculation methods compared in the study. Since the derivation of first-order finite difference techniques from Taylor series expansion is considered well-known, it is not described here.  Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

@@

It is possible to approximate derivatives quite accurately with a technique based on a complex-variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered for use in engineering analysis.  The basic idea is that a model parameter can be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} 
    f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} 
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}\left(h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im} \left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  
\label{eqn:complexFirstDeriv} 
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function. 
%
\todo[]{Re: II.6}For completeness, the forward and central difference formulas corresponding to equation~(\ref{eqn:complexFirstDeriv}) are \cite[Chap. 4.1.3]{chapra2010}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right), 
\label{eqn:forwardFirstDeriv} 
\end{equation}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). 
\label{eqn:centeredFirstDeriv} 
\end{equation} 
%
The order of the truncation error in the forward difference formula (equation~(\ref{eqn:forwardFirstDeriv})) indicates that for a given $h$ this method is less accurate an approximation than either central difference or complex-step which have errors of $\mathcal{O}(h^2)$.

\todo[]{Re: I.4}As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler \cite{lyness1967numerical}.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD \cite{squire1998using} in computing numerical derivatives. Papers by Newman et al.\ \cite{newman1998} and Anderson et al.\ \cite{anderson2001sensitivity} discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute tangent-stiffnesss of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ \cite{perez2000numerical, perez2012numerical} used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ \cite{martins2000automated} investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.

Martins et al.\ \cite{martins2003complex} demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis \cite{lai2008extensions} discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham \cite{al2010complex} investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ \cite{jin2010improved} discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ \cite{voorhees2011complex} discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex-numbers where required. In a follow-on work,  Millwater et al.\  \cite{millwater2013application} extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS \cite{systemes2012abaqus} FEA software package.  Nishida et al.\ \cite{nishida2013} used the complex-step method to compute tangent-stiffness matrices for implementation in a Levenberg-Marquardt optimization routine for design-of-experiments purposes.  Finally, Abreu et al.\ \cite{abreu2013generalization} discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  This recent literature has introduced the application of CS to single-dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, optimization, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  The current paper presents a comparison of the complex-step method to the other methods for numerical differentiation in as close to a one-to-one implementation as possible for direct comparison purposes in a large-scale engineering analysis code. To the author's knowledge it is the first application of the complex-step method to solve peridynamics mechanics problems, and the first implementation and comparison of the method in large-scale distributed memory parallel setting.

@@

An AD system evaluates composition functions in the expected manner: it works by first evaluating the innermost function of the composition, then presenting that function's output as input to the next level function until all levels are complete, observing commonly expected order of operations. However, an AD system does additional work during a function evaluation in that as each nested function is evaluated, the function's partial derivative with respect to the designated variables of the given input is also calculated.  This is possible because the elementary math functions are hard-coded into the AD source-code along with their analytical derivatives, and linked by special instructions, so that when the elementary math functions are called upon for computation, their partial derivatives may be computed and stored in a sequence. The AD system then multiplies the final sequence of partial derivatives together to produce the exact equivalent to taking a partial derivative of the corresponding composition function with respect to a designated variable at a particular value. 
%
It is obvious, but bears mentioning, that the AD system could simply store one value for partial derivatives, modifying it as appropriate for every function evaluation rather than keeping a sequence. This is significant because modifying a single value rather than keeping a list of values corresponding to every level of a composition function, which itself may not be the sole one needing to be evaluated, represents a savings in memory usage, which at large scales is an active concern. In the literature, the AD scheme described here is called \emph{forward automatic differentiation}. Only forward AD will be covered here since it is the implementation used in \emph{Sacado} and therefore in this study; however, the reader is referred to the introduction section of \cite{ref-AD-methods} and its citations for further information on AD, and
particularly \cite{ref-on-AD} which is foundational.

@@

While, for simplicity, the examples shown in the following refer only to linear problems, tangent-stiffness matrices generally arise in the context of non-linear analysis where Newton's method or quasi-Newton methods are used in the minimization of a given residual function.  In this context, the tangent-stiffness matrix can be thought of as the linearization of the system about a particular configuration.  It has been shown \cite{hughes1978consistent,hughes1978unconditionally} that in order to preserve the quadratic convergence properties of the Newton's methods, it is necessary that this linearization is carried out in a manner that is \emph{consistent} with the algorithmic constraint equations used when computing the internal forces that arise due to deformations.  An example of these constraint equations would be the Kuhn-Tucker conditions \cite{simo1998} that are used in integration of a flow rule for plasticity modeling.  If the continuum tangent-modulii are used in place of the \emph{consistent} or \emph{algorithmic tangent moduli} in the solution of a non-linear solid mechanics problem, convergence may still be achieved, but not in the quadratic manner that makes Newton's method attractive for this class of problems.  In the setting of non-linear solid mechanics, there is a very small class of material model algorithms in which consistent tangent-stiffness operators can be derived analytically.  Perfect plasticity and plasticity with isotropic hardening used in combination with a general nearest-point projection or \emph{radial return algorithm} are a few examples.  Therefore, when general models (and certainly more complex ones) are implemented into a general purpose computational mechanics code, the tangent-stiffness operators are typically defined via a numerical approximation.

@@

\section{\todo[]{Re: I.5}Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{subsec:Validation}

@@

\caption{Beam and spring simulated in example program.} 

@@

For this system, the equilibrium reaction force associated with a displacement $D$ of the spring is shown in equation~(\ref{eqn:TrussForce}) 

@@

    f(D) = \frac 1 2 AE(\cos^{2}\phi)(\frac{D}{L_{0}})^{2}[\frac{D}{L_{0}}\cos^{2}\phi - 3\sin\phi] + k_{s}D + (AE\frac{D}{L_{0}})\sin^{2}\phi,

@@

and the analytical tangent-stiffness is given by equation~(\ref{eqn:TrussStiffness})

@@

    S_{T}(D) = \frac 3 2 AE(\cos^{2}\phi)[\frac{D}{L_{0}}\cos^{2}\phi - 2sin\phi](\frac{D}{L_{0}^{2}}) + k_{s} + \frac{AE\sin^{2}\phi}{L_{0}}.

@@

where $\phi$ is a constant truss angle, $AE$ is the axial rigidity of the truss, $L_{0}$ is original truss member length, and $k_{s}$ is the Hooke's constant for the spring attached to the truss member. For a demonstration and verification of each of the methods as well as a convergence rate study, this truss and spring model was analyzed with each of the methods and compared with the analytic solution. As in the reference \cite{rezaiee2010dynamic}, equilibrium solutions for 12 incremental load steps of $\Delta P=\SI{4.448}{\newton}$ were found. Solver parameters common to all tests were a maximum limit of 100 iterations for convergence, a residual change tolerance of $\SI{1e-9}{\meter}$.  Data collected included number of iterations to converge summed over all load-steps and final displacement $D$, after the final load-step. Additionally finite difference step-size was varied in order to measure the impact on each of the approximate tangent-stiffness calculation methods. Naturally, varying step size does not affect AD or the analytically derived tangent-stiffness since those are not functions of step size. The accuracy results for verifying the tangent-stiffness calculation methods are summarized in Table~\ref{tab:ConvergenceStudy}. 

@@

  \caption{Displacement, D, after load-step 12.} \label{tab:Verification}   

@@

         Step exponent & Actual $\si{\centi\meter}$ & AD $\si{\centi\meter}$ & CS $\si{\centi\meter}$ & FD $\si{\centi\meter}$\\ 

@@

Table.~\ref{tab:Verification} shows that each of the methods solved the problem as accurately as one another within the precision reported. The exception to this is that the finite difference method failed due to an arithmetic error in association with extremely small step size, suggesting subtractive cancellation. However, since there is no accuracy reason for choosing an extremely small step size, each of the methods can be used to produce acceptable solutions to the nonlinear truss system. Table~\ref{tab:ConvergenceStudy} shows the total number of iterations taken over all load-steps for each of the methods. 

@@

         Step exponent & Actual $\si{\centi\meter}$ & AD $\si{\centi\meter}$ & CS $\si{\centi\meter}$ & FD $\si{\centi\meter}$\\ 

@@

Table~\ref{tab:ConvergenceStudy} indicates that when using AD or the `actual' analytically derived tangent-stiffness, a minimum number of iterations is achieved corresponding to a maximum convergence rate. CS and FD are similarly affected by a grossly large step-size, while subtractive cancellation invalidates the results for FD for very small step-sizes. Inaccuracy for large step-sizes is as predicted by the FD and CS formulas, which require a small step-size to allow the neglect of higher order terms. The results of this test show that step-size selection is not unimportant when using FD, but when properly configured any of the methods may return the same maximum convergence rate for this problem.  Overall conclusions regarding the methods that the accuracy of the tangent-stiffness should not be confused with the accuracy of the equilibrium solution, since each of the methods discussed here may be equally capable given enough iterations. Numerical measurements of accuracy besides number of iterations taken are not possible here because there is not basis for comparison, as the methods are solving the problem separately and producing different local Newton iterates. Additionally, while the inaccuracy of the tangent-stiffness may strongly affect convergence rates, this relationship was not precisely quantified here.

>>>>

%Package for adding notes/comments/and tracking changes
\usepackage[disable, color=red!40, textsize=scriptsize]{todonotes}

@@

    In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an underexplored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the  method with other techniques for numerical derivative calculation: automatic differentiation, forward finite difference, and central finite difference. The complex-step method was implemented in a massively parallel computational peridynamics code for the purpose of this comparison.  The methods were compared through profiling of the code for accuracy, speed, efficiency, and parallel scalability. This research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code.  Additionally, motivated by the reproducible research movement, all of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Newton's method \sep Newton-Raphson \sep numerical differentiation \sep complex-step \sep finite difference \sep automatic differentiation \sep tangent-stiffness
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%\MSC [2010] 65D25 
\end{keyword}

@@

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent\todo[]{Re: II.1}\footnote{In the context of plasticity theory, the algorithmic consistency of the tangent-stiffness matrix refers to consistency with the integration of a plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.} tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

@@

The scope of the application component of this study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the  \todo[]{Re: I.1} complex-step method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of forward difference, central difference, and automatic-differentiation methods needed for comparison to the complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as those required by this study.

The aim of this paper is not only to \todo[]{Re: I.1} study the complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving nonlinear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: In Section~\ref{subsec:TS} a description of tangent-stiffness matrices, detailed directions for producing tangent-stiffness matrices with each of the methods identified above, and a description and justification of the  \todo[]{Re: I.1} complex-step method for calculating tangent-stiffness matrices. Section~\ref{sec:Validation} provides a simple illustrative example for using each of the methods for calculating tangent-stiffness matrices in the solution of nonlinear problems. In Section~\ref{subsubsec:Impl} a description of implementing the complex-step method in the \emph{Peridigm} software. In Section~\ref{JGAM} a description of the quantities-of-interest used to rank the methods. In Sections~\ref{sec:Res} and \ref{sec:PeridigmConvergenceStudy} a presentation and analysis of the results of the comparative study, and in Section~\ref{sec:Conc} conclusions and thoughts on potential future work.  Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving nonlinear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.  The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion.  \todo[]{Re: II.5}This open source resource is available at the corresponding author's website or at \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

This section contains background information on the differentiation techniques underlying the tangent-stiffness matrix calculation methods compared in this study. Since the derivation of first-order finite difference techniques from Taylor series expansion is considered well-known, it is not described here.  Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

@@

It is possible to approximate derivatives quite accurately with a technique based on a complex-variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered for use in engineering analysis.  The basic idea is that a model parameter can be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} 
    f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} 
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}(h^2)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im} \left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  
\label{eqn:complexFirstDeriv} 
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function. 
%
\todo[]{Re: II.6}For completeness, the forward and central difference formulas corresponding to equation~(\ref{eqn:complexFirstDeriv}) are \cite[Chap. 4.1.3]{chapra2010}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right), 
\label{eqn:forwardFirstDeriv} 
\end{equation}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). 
\label{eqn:centeredFirstDeriv} 
\end{equation} 
%
The order of the truncation error in the forward difference formula (equation~(\ref{eqn:forwardFirstDeriv})) indicates that for a given $h$ this method is less accurate than either central difference or complex-step which have errors of $\mathcal{O}(h^2)$.

\todo[]{Re: I.4}As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler \cite{lyness1967numerical}.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original complex-step (CS) method to make it suitable for use on digital computers \cite{lyness1968differentiation}.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to finite-difference (FD) and central-difference (CD) \cite{squire1998using} in computing numerical derivatives. Papers by Newman et al.\ \cite{newman1998} and Anderson et al.\ \cite{anderson2001sensitivity} discuss computing sensitivities in a computational fluid dynamics (CFD) analysis using CS and compares the results to sensitivities produced with central difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute tangent-stiffnesss of residuals for use in Newton-Krylov schemes for solving nonlinear systems. Perez-Foguet et al.\ \cite{perez2000numerical, perez2012numerical} used the CS method to compute algorithmically consistent elastoplastic moduli in nonlinear solid mechanics problems.  A paper by Joaquim et al.\ \cite{martins2000automated} investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.

Martins et al.\ \cite{martins2003complex} demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis \cite{lai2008extensions} discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham \cite{al2010complex} investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ \cite{jin2010improved} discuss using CS to compute sensitivities in a solid mechanics finite element method (FEM) analysis.  Voorhees et al.\ \cite{voorhees2011complex} discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex-numbers where required. In a follow-on work,  Millwater et al.\  \cite{millwater2013application} extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS \cite{systemes2012abaqus} finite element analysis (FEA) software package.  Nishida et al.\ \cite{nishida2013} used the complex-step method to compute tangent-stiffness matrices for implementation in a Levenberg-Marquardt optimization routine for design-of-experiments purposes.  Finally, Abreu et al.\ \cite{abreu2013generalization} discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  This recent literature has introduced the application of CS to single-dimension and multidimensional derivatives, sensitivity analysis, nonlinear solution schemes, optimization, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  The current paper presents a comparison of the complex-step method to the other methods for numerical differentiation in as close to a one-to-one implementation as possible for direct comparison purposes in a large-scale engineering analysis code. To the author's knowledge it is the first application of the complex-step method to solve peridynamics mechanics problems, and the first implementation and comparison of the method in large-scale distributed memory parallel setting.

@@

An AD system evaluates a composition of functions in the expected manner: it works by first evaluating the innermost function of the composition, then presenting that function's output as input to the next level function until all levels are complete, observing commonly expected order of operations. However, an AD system does additional work during a function evaluation in that as each nested function is evaluated, the function's partial derivative with respect to the designated variables of the given input is also calculated.  This is possible because the elementary math functions are hard-coded into the AD source-code along with their analytical derivatives, and linked by special instructions, so that when the elementary math functions are called upon for computation, their partial derivatives may be computed and stored in a sequence. The AD system then multiplies the final sequence of partial derivatives together to produce the exact equivalent to taking a partial derivative of the corresponding composition function with respect to a designated variable at a particular value. 
%
It is obvious, but bears mentioning, that the AD system could simply store one value for partial derivatives, modifying it as appropriate for every function evaluation rather than keeping a sequence. This is significant because modifying a single value rather than keeping a list of values corresponding to every level of a composition function, which itself may not be the sole one needing to be evaluated, represents a savings in memory usage, which at large scales is an active concern. In the literature, the AD scheme described here is called \emph{forward automatic differentiation}. Only forward AD will be covered here since it is the implementation used in \emph{Sacado} and therefore in this study; however, the reader is referred to the introduction section of \cite{ref-AD-methods} and its citations for further information on AD, and
particularly \cite{ref-on-AD} which is foundational.

@@

While, for simplicity, the examples shown in the following refer only to linear problems, tangent-stiffness matrices generally arise in the context of nonlinear analysis where Newton's method or quasi-Newton methods are used in the minimization of a given residual function.  In this context, the tangent-stiffness matrix can be thought of as the linearization of the system about a particular configuration.  It has been shown \cite{hughes1978consistent,hughes1978unconditionally} that in order to preserve the quadratic convergence properties of the Newton's methods, it is necessary that this linearization is carried out in a manner that is \emph{consistent} with the algorithmic constraint equations used when computing the internal forces that arise due to deformations.  An example of these constraint equations would be the Kuhn-Tucker conditions \cite{simo1998} that are used in integration of a flow rule for plasticity modeling.  If the continuum tangent-modulii are used in place of the \emph{consistent} or \emph{algorithmic tangent moduli} in the solution of a nonlinear solid mechanics problem, convergence may still be achieved, but not in the quadratic manner that makes Newton's method attractive for this class of problems.  In the setting of nonlinear solid mechanics, there is a very small class of material model algorithms in which consistent tangent-stiffness operators can be derived analytically.  Perfect plasticity and plasticity with isotropic hardening used in combination with a general nearest-point projection or \emph{radial return algorithm} are a few examples.  Therefore, when general models (and certainly more complex ones) are implemented into a general purpose computational mechanics code, the tangent-stiffness operators are typically defined via a numerical approximation.

@@

\section{\todo[]{Re: I.5}Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{sec:Validation}

@@

\caption{Beam and spring simulated in the example program.} 

@@

For this system, the equilibrium reaction force associated with a displacement $D$ of the spring is 

@@

    f(D) = \frac 1 2 AE(\cos^{2}\phi)\left(\frac{D}{L_{0}}\right)^{2}\left[\frac{D}{L_{0}}\cos^{2}\phi - 3\sin\phi\right] + k_{s}D + \left(AE\frac{D}{L_{0}}\right)\sin^{2}\phi,

@@

and the analytical tangent-stiffness is 

@@

    S_{T}(D) = \frac 3 2 AE(\cos^{2}\phi)\left[\frac{D}{L_{0}}\cos^{2}\phi - 2sin\phi\right]\left(\frac{D}{L_{0}^{2}}\right) + k_{s} + \frac{AE\sin^{2}\phi}{L_{0}}.

@@

where $\phi$ is a constant truss angle, $AE$ is the axial rigidity of the truss, $L_{0}$ is original truss member length, $k_{s}$ is the Hooke's constant for the spring attached to the truss member, and $P$ is the applied load. For a demonstration and verification of each of the methods as well as a convergence rate study, this truss and spring model was analyzed with each of the methods and compared with the analytic solution. As in the reference \cite{rezaiee2010dynamic}, equilibrium solutions for 12 incremental load steps of $\Delta P=\SI{4.448}{\newton}$ were found. Solver parameters common to all tests were a maximum limit of 100 iterations for convergence, and a convergence tolerance for the change in displacement over an iteration $\Delta D \le \SI{1e-9}{\meter}$.  Data collected included number of iterations to converge summed over all load-steps and final displacement $D$, after the final load-step. Additionally finite difference and complex-step step sizes were varied in order to measure the impact on each of the approximate tangent-stiffness calculation methods. Naturally, varying the step size does not affect AD or the analytically derived tangent-stiffness since those are not functions of the step size. The accuracy results for verifying the tangent-stiffness calculation methods are summarized in Table~\ref{tab:ConvergenceStudy}. 

@@

  \caption{Displacement, D, after load-step 12} \label{tab:Verification}   

@@

         Step exponent & Analytical ($\si{\centi\meter}$) & AD ($\si{\centi\meter}$) & CS ($\si{\centi\meter}$) & FD ($\si{\centi\meter}$)\\ 

@@

The label ``Step Exponent'' refers to the value of $n$ in $h=10^{n}$ where $h$ is the finite or complex-step step size, and the symbol (") indicates that there is no change in the value from the value directly to its left.  Table.~\ref{tab:Verification} shows that each of the methods solved the problem as accurately as one another within the precision reported. The exception to this is that the finite difference method failed due to an arithmetic error in association with extremely small step size, suggesting subtractive cancellation. However, since there is no accuracy reason for choosing an extremely small step size, each of the methods can be used to produce acceptable solutions to the nonlinear truss system. Table~\ref{tab:ConvergenceStudy} shows the total number of iterations taken over all load-steps for each of the methods. 

@@

         Step exponent & Actual ($\si{\centi\meter}$) & AD ($\si{\centi\meter}$) & CS ($\si{\centi\meter}$) & FD ($\si{\centi\meter}$)\\ 

@@

Table~\ref{tab:ConvergenceStudy} indicates that when using AD or the `actual' analytically derived tangent-stiffness, a minimum number of iterations is achieved corresponding to a maximum convergence rate. CS and FD are similarly affected by a grossly large step-size, while subtractive cancellation invalidates the results for FD for very small step-sizes. Inaccuracy for large step-sizes is as predicted by the FD and CS formulas, which require a small step-size to allow the neglect of higher order terms. The results of this test show that step-size selection is not unimportant when using FD, but when properly configured any of the methods may return the same maximum convergence rate for this problem.  Conclusions regarding the methods that the accuracy of the tangent-stiffness should not be confused with the accuracy of the equilibrium solution, since each of the methods discussed here may be equally capable given enough iterations. Numerical measurements of accuracy besides number of iterations taken are not possible here because there is not basis for comparison, as the methods are solving the problem separately and producing different local Newton iterates. Additionally, while the inaccuracy of the tangent-stiffness may strongly affect convergence rates, this relationship was not precisely quantified here.
